---
title: "Project Milestone #2"
author: "Group 10: Ruby Yun, Jackie Zintel"
date: "2026-09-30"
output: html_document
---

Final Milestone #2

**1. Description of dataset**
- We used scenario_1 datasets from the PHW251 Project Data GitHub repository.
- The data is simulated, and the population estimates from from the CA Dept of Finance for the year 2023.
- Research Question: How did infection rates during the simulated outbreak vary by county, age, sex, and race across California?
- Our dataset contains simulated infectious disease outbreak data across CA, including number of new infections as well as demographic and geographic characteristics.These variables allow us to examine how the burden of infection differed across populations and locations during the outbreak. The dataset also includes 2023 CA Dept of Finance population estimates, which provides the denominators needed to calculate infection rates rather than simply comparing case counts.

**2. Import dataset**

```{r}
library(tidyverse)
library(janitor)

#import datasets
getwd()
ca_pop <- read.csv("../phw251_projectdata/scenario_1/ca_pop_2023.csv")
head(ca_pop)

sim_ca <- read.csv("../phw251_projectdata/scenario_1/sim_novelid_CA.csv")
head(sim_ca)

sim_la <- read.csv("../phw251_projectdata/scenario_1/sim_novelid_LACounty.csv")
head(sim_la)

#clean the column names
ca_pop <- ca_pop %>%
  clean_names()
sim_ca <- sim_ca %>%
  clean_names()
sim_la <- sim_la %>%
  clean_names()

```

\newpage

**3. Data types and variables**

```{r}

# Identify all the variables in each dataset
sapply(ca_pop, class)
sapply(sim_ca, class)
sapply(sim_la, class)

```

- We chose the following variables: county, age_cat, sex, race_ethnicity, new_infections, pop

- Desired data types for analysis
  - County, age category, sex, and race: character/categorical
  - Population and infection/severity counts: integer/numeric
  - Diagnosis/report dates: Date
  - Time/week variables: integer

- Variables that need conversion/cleaning:
  - sim_ca$race_ethnicity: currently integer codes (1, 2, 3, 4, etc.)
  - Need to recode as character race categories that match ca_pop$race7 (e.g., "Black NH").
  - sim_ca$dt_diagnosis: character needs to be converted to Date

\newpage

**4. Descriptive Statistics**

- Basic description provided for all 5+ variables
  - county: County for population
  - age_cat: Age Category
  - sex: Sex 
  - race_ethnicity: Race-ethnicity categorization as defined by CA Dept of Finance 
1 = "White, Non-Hispanic", 2 = "Black, Non-Hispanic", 3 = "American Indian or Alaska Native, Non-Hispanic", 4 = "Asian, Non-Hispanic", 5 = "Native Hawaiian or Pacific Islander, Non-Hispanic", 6 = "Multiracial (two or more of above races),       Non-Hispanic", 7 = "Hispanic (any race)", 9 = "Unknown)
  - new_infections: Number of newly diagnosed individuals
  - pop: CA Dept of Finance estimated all ages population for year 2023

- Numeric variables: mean, median, range

```{r numeric variables}
paste("Mean, median, and range of California population")
ca_numeric <- c(mean = mean(ca_pop$pop, na.rm = T),
                median = median(ca_pop$pop, na.rm = T),
                min = min(ca_pop$pop, na.rm = T),
                max = max(ca_pop$pop, na.rm = T))
round(ca_numeric, 2)

paste("Mean, median, and range of new infections")
sim_ca_numeric <- c(mean = mean(sim_ca$new_infections, na.rm = T),
                median = median(sim_ca$new_infections, na.rm = T),
                min = min(sim_ca$new_infections, na.rm = T),
                max = max(sim_ca$new_infections, na.rm = T))
round(sim_ca_numeric, 2)

```

- Character/categorical variables: unique values/categories

```{r chr/cat variables}
unique(ca_pop$county)
unique(ca_pop$age_cat)
unique(ca_pop$sex)
unique(sim_ca$race_ethnicity)
```

\newpage

**5. Data Cleaning**

```{r join the datasets}



```

- Variables that need to be converted (e.g., chr -> numeric)
- Values that need to be cleaned (e.g., Male, male, M)
- Re-categorization
- Missing values identified (e.g., NA, Unknown, N/A, etc.)
- Variables that may be useful for future joins (e.g., geographic variables)

\newpage

**6. Final Submission**
- Rmd or Qmd completed
- R code & output included
- Rendered to HTML
- HTML uploaded/submitted through GitHub
